Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-09-06 concurrency/CPP 2024/paper.tex
first attempt at precisely defining a template

@@

Our specific contributions are:
\begin{itemize}
	\item We reimplement the template approach without flow interfaces, with a simple interface involving only the concept of ``keys belonging in this node/subtree''.
	\item We implement the template approach in VST, allowing us to apply it to C programs and obtain end-to-end correctness proofs in a single verification system.
	\item To the best of our knowledge, this is the first mechanized verification of a template approach to concurrent data structure implementations in a real-world programming language, and its first application to data structures not written specifically as case studies.
	\item We highlight several questions about the boundaries of the template/data structure division, and describe a limitation of the template approach on tree data structures.
\end{itemize}

@@

Logical atomicity is a separation logic technique for concisely specifying the behavior of a concurrent operation. A logically atomic triple has the form $\forall a.\ \left\langle \texttt{P}_l\ |\ \texttt{P}_p(a) \right\rangle\ \texttt{c}\ \left\langle \texttt{Q}_l\ |\ \texttt{Q}_p(a)\right\rangle$, where $\texttt{P}_l$ and $\texttt{Q}_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $\texttt{P}_p$ and $\texttt{Q}_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure. Intuitively, this says that \lstinline{c} is an operation on an abstract object (i.e., data structure) $a$, and its effect is to \emph{atomically} transform $a$ from a state satisfying $\texttt{P}_p$ to a state satisfying $\texttt{Q}_p$, with no intermediate states visible to any other thread. More precisely, the triple asserts that if $\texttt{P}_l$ holds true before a call to \lstinline{c} and $\texttt{P}_p$ is true for some value of $a$ in a shared state, then $\texttt{P}_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $\texttt{c}$, at which point $\texttt{Q}_p$ will become true atomically for the same value $a$ (and $\texttt{Q}_l$ will be true after $\texttt{c}$ ends).

As an example, a sequential stack \lstinline{push} operation could conventionally be specified as 
$\{\mathsf{stack}\ s\ p\}\ \texttt{push}(v)\ \{\mathsf{stack}\ (v :: s)\ p\}$. Its concurrent counterpart could be specified as 
\begin{mathpar}
	\forall s.\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ s\right\rangle\ \\ \texttt{push}(v)\  \\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ (v::s)\right\rangle
\end{mathpar}
indicating that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $s$ to $v::s$ at some point during its execution. The stack itself is a shared resource, and can only be accessed and modified atomically by threads holding the corresponding \texttt{is\_stack} assertion. The local and shared assertions are connected by an arbitrary identifier $g$, which we will generally omit when clear from context. Logical atomicity can be used to prove linearizability---if all of a data structure's operations satisfy logically atomic triples derived from the corresponding sequential operations, the data structure is linearizable~\cite{la-lin}---but can also be used to specify more complex behavior of nonlinearizable data structures~\cite{compass}.

@@

@@

%\wm{The fonts are still off here---for instance, p, n, and n' are mathematical variables, not references to the C code. Fix if we have time.}
The \lstinline{traverse} function can then be specified as follows:
%\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.} 
\begin{mathpar}
	% spec of inRange
	{\color{blue}
		\forall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
		\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ m \
		\right\rangle
	}\vspace{-6pt}\\
	\vspace{-6pt}\texttt{traverse(pn, k)} \\
	{\color{blue}
		\left\langle \mathit{res}. \ \exists \  n', v.
		\begin{array}{l} \texttt{pn} \mapsto \left(n', n'\right) \ \ast \\ \nodeboxrep\left(n'\right) \ \ast\ \texttt{k} \in \mathsf{range}\ \ast \ 
			\\ 
			(\mathsf{if} \ \mathit{res} \ \\ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
			\\ \mathsf{else} \ \mathsf{node\_contents}\left(n', (\texttt{k}, v), \mathsf{range}\right))
		\end{array}
		\ \Bigg | \ \treerep\ m \
		\right\rangle
	}
\end{mathpar}
The local precondition of \lstinline{traverse} includes both the node handle $\nodeboxrep(n)$ and its contents $\mathsf{R}(n)$, indicating that a thread should already hold $n$'s lock before calling \lstinline{traverse}. The output of \lstinline{traverse} is a new node $n'$ in the data structure such that the key \texttt{k} falls within the range of $n'$. The local postcondition then includes the handle of the new node $\nodeboxrep(n')$, its contents $\mathsf{R}(n')$, and a Boolean variable $\mathit{res}$ indicating whether \lstinline{traverse} found an empty node or a node with key \lstinline{k}.

The resources $\mathsf{R}$ contained in a node depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node. Formally, the lock invariant is defined by 
\begin{align*}\mathsf{node\_contents}(n, c, \mathsf{range}) \triangleq \\ 
	\mathsf{ghost\_node}(n, c,\mathsf{range}) \ \ast \ \mathsf{node\_data}(n, c) \\
	\mathsf{R}(n) \triangleq \exists\ c, \mathsf{range}.\ \mathsf{node\_contents}(n, c, \mathsf{range})\end{align*}
where the definition of $\mathsf{node\_data}$ is supplied by the target data structure. The contents $c$ of a node can be either a key-value pair $(k, v)$, or the empty contents $\cdot$ (used for nodes that have been allocated but not yet assigned keys).

@@

\begin{figure*}[!ht]
	$\color{blue}
	\forall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
	\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ m \
	\right\rangle$
	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  n'.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast\ 
		\\ 
		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
	\end{array}
	\ \Bigg| \ \treerep\ m \
	\right\rangle$
	\caption{Proof outline of the lock-coupling \texttt{traverse} function}
	\label{proof_lock_traverse}
\end{figure*}

@@

The \lstinline{insert} function uses this specification of \lstinline{traverse} to update the state of the data structure. The specification of \texttt{insert} is:
\begin{mathpar}
	{\color{blue}
		m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\\ \texttt{insert(r, k, v)}\  \\
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
\end{mathpar}
where $m$ and $m[\texttt{k} \mapsto \texttt{v}]$ denote the tree's abstract states before and after the function's execution, respectively. The proof of \lstinline{insert} is outlined in Appendix \ref{sec:apd_proof}, but informally it is quite simple. We begin by initializing the \lstinline{pn} struct with the root node and acquiring its lock, allowing us to satisfy the precondition of \lstinline{traverse}. If \lstinline{traverse} returns 0, we have found a node with key \lstinline{k}, and all we need to do is update that node's value to \lstinline{v} (line 6 in Figure \ref{insert_lock}). Otherwise, \texttt{traverse} has reached an empty node with \lstinline{k} in its range. We then call the data-structure-specific \texttt{insertOp} function, which inserts the new key-value pair at the empty node. In either case, before we release the node's lock, we must show that we have altered the tree from its current abstract state $m$ to $m[\texttt{k} \mapsto \texttt{v}]$, thereby satisfying the public postcondition of \lstinline{insert}. We do this by demonstrating that in both cases, the update to the concrete data structure corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the data structure.
Finally, we release the lock acquired by \lstinline{traverse} and deallocate the \lstinline{pn} structure.

@@

The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
\begin{mathpar}
	{\color{blue}
		\forall \  m. \left\langle
		\begin{array}{c}
			\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ m
		\end{array}
		\right\rangle
	}
	\vspace{-6pt}\\ \vspace{-6pt}
	\texttt{traverse(pn, k)} 
	\\
	{\color{blue}
		\left\langle \mathit{res}. \ \exists \  n', v.
		\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ \\ \texttt{k} \in \mathsf{range}\ \ast\ 
			\\ 
			(\mathsf{if} \ \mathit{res} \\ 
			\ \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
			\\ 
			\ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range}))
		\end{array}
		\Bigg| \ \treerep\ m \
		\right\rangle
	}
\end{mathpar}
As before, the \texttt{traverse} function will navigate the structure and return a node $n'$ whose range includes \texttt{k}, as well as acquiring $n'$'s lock and returning its contents. Also as before, the Boolean $\mathit{res}$ indicates whether $n'$ is empty or contains the key \lstinline{k}.

@@

\section{Verifying Binary Search Trees using Templates}
\label{BST_proof}
In this section, we demonstrate how to instantiate the templates with specific data structures, giving us verified C implementations of concurrent data structures. The top-level specifications we obtain for search structure operations are shown below: each operation takes effect on the data structure atomically, modifying it or returning information as appropriate\footnote{Krishna et al.~\cite{templates} combine these specifications into a single function parameterized by the choice of operation; however, clients will still ultimately want to call \lstinline{insert}, \lstinline{lookup}, or \lstinline{delete} functions.}.
\begin{figure*}[!ht]
	\centering
\begin{mathpar}
	{\color{blue}
		\forall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{insert(r, k, v)}\ 
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
	\\
	{\color{blue}
		\forall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{lookup(r, k)}\ 
	{\color{blue}
		\left\langle \texttt{v}. \ \
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m \land m(\texttt{k}) = \texttt{v}
		\right\rangle
	}
	\\
	{\color{blue}
		\forall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{delete(r, k)}\ 
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \_ ])
		\right\rangle
	}
\end{mathpar}
\end{figure*}
As we saw in the previous section, these top-level specifications have already been verified at the template level; all we need to do now is provide implementations of data-structure-specific functions (e.g. \lstinline{findNext}) that satisfy the right specifications to instantiate the template proofs. Once we have chosen a template and a data structure, we should not need to do any further reasoning about specific concurrency patterns.

@@

\begin{figure*}[!ht]
	$\color{blue}
	\forall \  m. \left\langle
	\begin{array}{c}
		\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ m
	\end{array}
	\right\rangle$
	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  n'.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ \texttt{k} \in \mathsf{range}\ \ast \ 
		\\ 
		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
	\end{array}
	\ \Bigg| \ \treerep\ m \
	\right\rangle$
	\caption{Proof outline of the give-up \texttt{traverse} function}
	\label{proof_giveup_traverse}
\end{figure*}

\begin{figure*}[!ht]
	\begin{subfigure}{\textwidth}
		$\color{blue}
		\forall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_insert.c}
		$\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle$
		\caption{Proof outline of the lock-coupling \texttt{insert} function}
		\label{proof_lock_insert}	
	\end{subfigure}\qquad
	\begin{subfigure}{\textwidth}
		$\color{blue}
		\forall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_insert.c}
		$\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle$
		\caption{Proof outline of the give-up \texttt{insert} function}
		\label{proof_giveup_insert}
	\end{subfigure}
	\caption{Proof outlines for the \texttt{insert} function}
	\label{proof_lock_giveup_insert}
\end{figure*}

>>>>

%https://tex.stackexchange.com/questions/78187/is-there-another-symbol-that-is-slightly-different-from-forall-likewise-for-e
\makeatletter
\newcommand*{\fforall}{%
  {\mathpalette\fforallAux{}}%
}
\newcommand*{\fforallAuxx}[1]{%
  \sbox0{$\m@th#1\forall$}%
  \sbox2{%
    \rlap{%
      \raisebox{\depth}{$\m@th#1\backslash$}%
    }%
    \kern\ht0 %
  }%
  \sbox2{\resizebox{\ht2}{\height}{\copy2}}%
  \sbox2{\resizebox{!}{\ht0}{\copy2}}%
  \wd2=0pt %
  \copy2
  \forall
}
\newsavebox\forallBox
\newdimen\forallLineWidth
\newdimen\forallSep
\newcommand*{\fforallAux}[1]{%
  \sbox\forallBox{$\m@th#1\forall$}%
  \setlength{\forallLineWidth}{.06\wd\forallBox}%
  \setlength{\forallSep}{.09\wd\forallBox}%
  \tikz[
    inner sep=0pt,
    line cap=round,
    line width=\forallLineWidth,
  ]
  \draw
    (0,0) node (A) {\copy\forallBox}
    (A.south) ++(-\forallSep-\forallLineWidth,.4\forallLineWidth)
    coordinate (A1)
    (A.north west) ++(-\forallSep,-\forallLineWidth)
    coordinate (A2)
    (A1) -- (A2)
  ;%
}
\makeatother

@@

Our specific contributions are:
\begin{itemize}
	\item We reimplement the template approach without flow interfaces, with a simple interface involving only the concept of ``keys belonging in this node/subtree''.
	\item We implement the template approach in VST, allowing us to apply it to C programs and obtain end-to-end correctness proofs in a single verification system.
	\item To the best of our knowledge, this is the first mechanized verification of a template approach to concurrent data structure implementations in a real-world programming language, and its first application to data structures not written specifically as case studies.
	\item We highlight several questions about the boundaries of the template/data structure division, and clarify an element of the template approach that has not been formally addressed.
\end{itemize}

@@

Logical atomicity is a separation logic technique for concisely specifying the behavior of a concurrent operation. A logically atomic triple has the form $\fforall a.\ \left\langle \texttt{P}_l\ |\ \texttt{P}_p(a) \right\rangle\ \texttt{c}\ \left\langle \texttt{Q}_l\ |\ \texttt{Q}_p(a)\right\rangle$, where $\texttt{P}_l$ and $\texttt{Q}_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $\texttt{P}_p$ and $\texttt{Q}_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure. Intuitively, this says that \lstinline{c} is an operation on an abstract object (i.e., data structure) $a$, and its effect is to \emph{atomically} transform $a$ from a state satisfying $\texttt{P}_p$ to a state satisfying $\texttt{Q}_p$, with no intermediate states visible to any other thread. More precisely, the triple asserts that if $\texttt{P}_l$ holds true before a call to \lstinline{c} and $\texttt{P}_p$ is true for some value of $a$ in a shared state, then $\texttt{P}_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $\texttt{c}$, at which point $\texttt{Q}_p$ will become true atomically for the same value $a$ (and $\texttt{Q}_l$ will be true after $\texttt{c}$ ends).

As an example, a sequential stack \lstinline{push} operation could conventionally be specified as 
$\{\mathsf{stack}\ s\ p\}\ \texttt{push}(v)\ \{\mathsf{stack}\ (v :: s)\ p\}$. Its concurrent counterpart could be specified as 
\begin{mathpar}
	\fforall s.\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ s\right\rangle\ \\ \texttt{push}(v)\  \\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ (v::s)\right\rangle
\end{mathpar}
indicating that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $s$ to $v::s$ at some point during its execution. The stack itself is a shared resource, and can only be accessed and modified atomically by threads holding the corresponding \texttt{is\_stack} assertion. The local and shared assertions are connected by an arbitrary identifier $g$, which we will generally omit when clear from context. Logical atomicity can be used to prove linearizability---if all of a data structure's operations satisfy logically atomic triples derived from the corresponding sequential operations, the data structure is linearizable~\cite{la-lin}---but can also be used to specify more complex behavior of nonlinearizable data structures~\cite{compass}.

@@

\subsection{What is a Search Structure Template?}
Krishna et al.~\cite{templates} described the search structure template approach and used it to verify (parts of) several concurrent data structures. Their examples strongly suggest a systematic approach to separating sequential data structure logic from concurrent synchronization patterns. However, the approach is not formally defined, and on closer inspection its inputs and outputs are not identical across the example templates. In this section, we attempt to precisely describe the pieces of the search structure template approach, and how we can know whether we have successfully verified a data structure given a collection of verified sequential and concurrent functions.

A concurrent search structure is a data structure that supports three operations: \lstinline{insert}, \lstinline{lookup}, and \lstinline{delete}. It is intended to implement a map from keys to values, and to behave correctly when accessed simultaneously by any number of threads. In separation logic, we can prove correctness of a search structure by showing that its operations satisfy the logically atomic specifications shown in Figure~\ref{search-specs}. In other words, each operation takes effect atomically on an abstract map $m$, reading and/or updating it as appropriate to the operation.

\begin{figure*}[!ht]
	\centering
\begin{mathpar}
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{insert(r, k, v)}\ 
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
	\\
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{lookup(r, k)}\ 
	{\color{blue}
		\left\langle \texttt{v}. \ \
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m \land m(\texttt{k}) = \texttt{v}
		\right\rangle
	}
	\\
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{delete(r, k)}\ 
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \_ ])
		\right\rangle
	}
\end{mathpar}
\caption{The logically atomic specifications for search structure operations}
\label{search-specs}
\end{figure*}

A search structure template is an implementation of these three functions, written in a particularly restricted way. Aside from some top-level management code (e.g., acquiring and releasing an initial lock), all that each function should do is call two functions: a function called \lstinline{traverse}, which is specific to the template and implements its synchronization mechanism, and a concurrency-unaware function (e.g. \lstinline{insertOp}) provided by the target data structure. (In fact, Krishna et al. implemented all three operations as a single function \lstinline{decisiveOp}, parameterized by a function that takes the target operation type as input and performs the appropriate data structure operation; for ease of adaptation to C, we write three different functions, but their structures are identical.) A template consists of an implementation of the top-level operations and the \lstinline{traverse} function used to define them; a data structure consists of an implementation of \lstinline{insertOp}, etc., as well as a function \lstinline{findNext} that is used to implement \lstinline{traverse}. The interface between them is a set of standard specifications for \lstinline{insertOp}, etc., and \lstinline{findNext}: if a template's functions can be verified assuming these standard specifications, and a data structure's implementations can be shown to meet those specifications, then the template and the data structure can be combined to yield a verified concurrent search structure.

This is the theory behind the template approach to verification. Unfortunately, it is not realized in practice. In the Coq development of Krishna et al., different templates use slightly different specifications for data structure operations and \lstinline{findNext}; the specifications proved for data structure implementations do not always match those used by templates (which is possible because they use separate provers for the sequential and concurrent proofs); and some data structures rely on other functions that do not appear in the templates (e.g., maintenance operations that split full nodes in B-trees). In the following presentation, we follow the template plan as far as possible, noting the places where we must deviate from the uniform approach; we discuss maintenance operations in further detail in Section~\ref{internal_reorganize}.

@@

%\wm{The fonts are still off here---for instance, p, n, and n' are mathematical variables, not references to the C code. Fix if we have time.}
The \lstinline{traverse} function can then be specified as follows:
%\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.} 
\begin{mathpar}
	% spec of inRange
	{\color{blue}
		\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
		\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ m \
		\right\rangle
	}\vspace{-6pt}\\
	\vspace{-6pt}\texttt{traverse(pn, k)} \\
	{\color{blue}
		\left\langle \mathit{res}. \ \exists \  n', v.
		\begin{array}{l} \texttt{pn} \mapsto \left(n', n'\right) \ \ast \\ \nodeboxrep\left(n'\right) \ \ast\ \texttt{k} \in \mathsf{range}\ \ast \ 
			\\ 
			(\mathsf{if} \ \mathit{res} \ \\ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
			\\ \mathsf{else} \ \mathsf{node\_contents}\left(n', (\texttt{k}, v), \mathsf{range}\right))
		\end{array}
		\ \Bigg | \ \treerep\ m \
		\right\rangle
	}
\end{mathpar}
The local precondition of \lstinline{traverse} includes both the node handle $\nodeboxrep(n)$ and its contents $\mathsf{R}(n)$, indicating that a thread should already hold $n$'s lock before calling \lstinline{traverse}. The output of \lstinline{traverse} is a new node $n'$ in the data structure such that the key \texttt{k} falls within the range of $n'$. The local postcondition then includes the handle of the new node $\nodeboxrep(n')$, its contents $\mathsf{R}(n')$, and a Boolean variable $\mathit{res}$ indicating whether \lstinline{traverse} found an empty node or a node with key \lstinline{k}.

The resources $\mathsf{R}$ contained in a node depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node. Formally, the lock invariant is defined by 
\begin{align*}&\mathsf{node\_contents}(n, c, \mathsf{range}) \triangleq \\ 
	&\qquad\mathsf{ghost\_node}(n, c,\mathsf{range}) \ \ast \ \mathsf{node\_data}(n, c) \\
	&\mathsf{R}(n) \triangleq \exists\ c, \mathsf{range}.\ \mathsf{node\_contents}(n, c, \mathsf{range})\end{align*}
where the definition of $\mathsf{node\_data}$ is supplied by the target data structure. The contents $c$ of a node can be either a key-value pair $(k, v)$, or the empty contents $\cdot$ (used for nodes that have been allocated but not yet assigned keys).

@@

\begin{figure*}[!ht]
	$\color{blue}
	\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
	\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ m \
	\right\rangle$
	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  n'.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast\ 
		\\ 
		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
	\end{array}
	\ \Bigg| \ \treerep\ m \
	\right\rangle$
	\caption{Proof outline of the lock-coupling \texttt{traverse} function}
	\label{proof_lock_traverse}
\end{figure*}

@@

The \lstinline{insert} function uses this specification of \lstinline{traverse} to update the state of the data structure. The specification of \texttt{insert} is:
\begin{mathpar}
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\\ \texttt{insert(r, k, v)}\  \\
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
\end{mathpar}
where $m$ and $m[\texttt{k} \mapsto \texttt{v}]$ denote the tree's abstract states before and after the function's execution, respectively. The proof of \lstinline{insert} is outlined in Appendix \ref{sec:apd_proof}, but informally it is quite simple. We begin by initializing the \lstinline{pn} struct with the root node and acquiring its lock, allowing us to satisfy the precondition of \lstinline{traverse}. If \lstinline{traverse} returns 0, we have found a node with key \lstinline{k}, and all we need to do is update that node's value to \lstinline{v} (line 6 in Figure \ref{insert_lock}). Otherwise, \texttt{traverse} has reached an empty node with \lstinline{k} in its range. We then call the data-structure-specific \texttt{insertOp} function, which inserts the new key-value pair at the empty node. In either case, before we release the node's lock, we must show that we have altered the tree from its current abstract state $m$ to $m[\texttt{k} \mapsto \texttt{v}]$, thereby satisfying the public postcondition of \lstinline{insert}. We do this by demonstrating that in both cases, the update to the concrete data structure corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the data structure.
Finally, we release the lock acquired by \lstinline{traverse} and deallocate the \lstinline{pn} structure.

@@

The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
\begin{mathpar}
	{\color{blue}
		\fforall \  m. \left\langle
		\begin{array}{c}
			\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ m
		\end{array}
		\right\rangle
	}
	\vspace{-6pt}\\ \vspace{-6pt}
	\texttt{traverse(pn, k)} 
	\\
	{\color{blue}
		\left\langle \mathit{res}. \ \exists \  n', v.
		\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ \\ \texttt{k} \in \mathsf{range}\ \ast\ 
			\\ 
			(\mathsf{if} \ \mathit{res} \\ 
			\ \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
			\\ 
			\ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range}))
		\end{array}
		\Bigg| \ \treerep\ m \
		\right\rangle
	}
\end{mathpar}
As before, the \texttt{traverse} function will navigate the structure and return a node $n'$ whose range includes \texttt{k}, as well as acquiring $n'$'s lock and returning its contents. Also as before, the Boolean $\mathit{res}$ indicates whether $n'$ is empty or contains the key \lstinline{k}.

@@

\section{Verifying Binary Search Trees using Templates}
\label{BST_proof}
In this section, we demonstrate how to instantiate the templates with specific data structures, giving us verified C implementations of concurrent data structures. 
As we saw in the previous section, the top-level specifications of data structure operations have already been verified at the template level; all we need to do now is provide implementations of data-structure-specific functions (e.g. \lstinline{findNext}) that satisfy the right specifications to instantiate the template proofs. Once we have chosen a template and a data structure, we should not need to do any further reasoning about specific concurrency patterns.

@@

\begin{figure*}[!ht]
	$\color{blue}
	\fforall \  m. \left\langle
	\begin{array}{c}
		\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ m
	\end{array}
	\right\rangle$
	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  n'.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ \texttt{k} \in \mathsf{range}\ \ast \ 
		\\ 
		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
	\end{array}
	\ \Bigg| \ \treerep\ m \
	\right\rangle$
	\caption{Proof outline of the give-up \texttt{traverse} function}
	\label{proof_giveup_traverse}
\end{figure*}

\begin{figure*}[!ht]
	\begin{subfigure}{\textwidth}
		$\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_insert.c}
		$\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle$
		\caption{Proof outline of the lock-coupling \texttt{insert} function}
		\label{proof_lock_insert}	
	\end{subfigure}\qquad
	\begin{subfigure}{\textwidth}
		$\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_insert.c}
		$\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle$
		\caption{Proof outline of the give-up \texttt{insert} function}
		\label{proof_giveup_insert}
	\end{subfigure}
	\caption{Proof outlines for the \texttt{insert} function}
	\label{proof_lock_giveup_insert}
\end{figure*}
